\documentclass{scrartcl}
\usepackage[utf8]{inputenc}
\usepackage{hyperref}
\usepackage{url}
\usepackage{natbib}
\usepackage{graphicx}
\usepackage{cleveref} % this must be the last package to be loaded

\newcommand{\emailaddr}[1]{\href{mailto:#1}{\texttt{#1}}}

\title{\LARGE
    Final Report Template\\
    DS-Monopoly
}
\subtitle{Final Report for the Distributed Systems Course [2025/2026]}

% Consider watching:
% https://www.youtube.com/watch?v=ihxSUsJB_14
% https://www.youtube.com/watch?v=XTFWaV55uDo

\author{
    Daniele Loverre \\ \emailaddr{daniele.loverre@studio.unibo.it}
}

\date{\today}

\begin{document}
\nocite{*}

\maketitle

\begin{abstract}
  DS-Monopoly is a distributed clone of the Monopoly board game.
    It is a simplified version of the game where a centralized server acts as the \textit{bank},
  it coordinates the game and keeps the game state, validates actions, prevents cheating and
  broadcasts the updates to all the connected clients.
  The clients are \textit{players} that can join a game by connecting to the central server (2 to 4 players per game).
  Players can buy properties or houses, roll dice and see the board and the players information from a GUI.
  Disconnections from both sides, clients and server, are handled in order to keep the game running.
  Features like checkpointing, authentication and timeouts, are used to enhance fault tolerance and
  ensure more reliability and availability.
\end{abstract}

\section*{Disclaimer}

During the preparation of this work, the author used Gemini to assist in coding the graphical user interface
with PyGame for the clients and to accelerate the development of some small functions and code refactoring.
It was not used to design the project's architecture nor to write the final report.

After using this tool/service, 
the author reviewed and edited the content as needed
and takes full responsibility for the content of the final report/artifact.

\section{Concept}\label{concept}

The project is a desktop application with a GUI for clients.
Players can see the board with houses built and pawns on it on the left,
and the information of the players, such as their balance and their properties, on the right.
The last event of the game is also written on the GUI in order to
let players understand what happens during their and others player's turn.
The server can be started on CLI or using docker compose, all the information and logs of the server can be read on CLI.


Players cannot play on the same device (more precisely on the same window).
They connect to a server, providing the server's port and ip,
provided on the server's CLI at the start, and an available nickname.
Clients send messages every event they generate, created on button's pressing.
The server broadcasts to all his connected clients the game state on every update on it.
They exchange messages in JSON sent via TCP socket.
The only data stored are the checkpoint of the game state saved in a file on the server side
and the tokens saved in files for the authentication, one for each nickname used, on the client side.


When the server is started the game waits for clients to connect.
Once all the players in the \textit{lobby} are ready (at least 2 players)
the game starts, 5000 coins and 5 random properties are assigned to each player.
Following their turn, players can make actions that are sent to the server and updates the game state.
The updated state is sent back to all the players.
The game continues until only one players remains without going in \textit{bankruptcy}
or until all the players left the game.


Distribution is needed in this project to allow geographical distribution between nodes,
so that is not necessary that players are in the same physical room, but the game state is shared across the network.
The state is kept only on the server that acts as a single source of truth, avoiding cheating from clients.
Furthermore, the failure of a client does not interrupt the game, but it can be continued by the other players connected.

\section{Requirements Elicitation and Analysis}\label{requirements}
\subsection{Functional requirements}\label{functional_requirements}

\begin{itemize}
    \item \textbf{Server Initialization:} The server should expose and print its endpoint (IP and port) upon startup.
    \begin{itemize}
        \item \textit{Acceptance Criterion:} Running the server application outputs its endpoint on the CLI.
    \end{itemize}

    \item \textbf{Client Connection:} Clients should be able to connect to the server using the provided endpoint and an available nickname, opening a GUI lobby.
    \begin{itemize}
        \item \textit{Acceptance Criterion:} A user entering a valid IP, port, and unique nickname successfully joins the lobby and sees the GUI.
    \end{itemize}

    \item \textbf{Game Start Condition:} The game should start only when at least two players are in the lobby and all connected players pressed \textit{"ready"}.
    \begin{itemize}
        \item \textit{Acceptance Criterion:} The game goes from lobby to playing state immediately after the last player clicks the ready button.
    \end{itemize}

    \item \textbf{Turn Management:} The system should have a turn-based logic, each player can only do the actions they are allowed to do during their specific turn.
    \begin{itemize}
        \item \textit{Acceptance Criterion:} A player cannot click \textit{"roll dice"} or \textit{"buy"} if it is not their turn or if the action is invalid for their current state.
    \end{itemize}

    \item \textbf{Game State Visualization:} The GUI should display the game board, player information (balance, properties), and the last game event.
    \begin{itemize}
        \item \textit{Acceptance Criterion:} Whenever an action is performed, the GUI immediately updates the board, the balances, and the event log for all connected players.
    \end{itemize}

    \item \textbf{Client Disconnection Handling:} If a client disconnects, the game should pause and grant the disconnected player 30 seconds to rejoin.
    \begin{itemize}
        \item \textit{Acceptance Criterion:} A disconnection triggers a 30-second timer; if the player reconnects in time, the game resume. Otherwise, the player goes bankrupt.
    \end{itemize}

    \item \textbf{Server Disconnection Handling:} If the server fails, clients should wait up to 90 seconds for it to restart before shutting down.
    \begin{itemize}
        \item \textit{Acceptance Criterion:} Shutting down the server triggers a 90-second waiting state on all active clients.
    \end{itemize}

    \item \textbf{Winning Condition:} The game should end when only one player remains active and not in bankruptcy.
    \begin{itemize}
        \item \textit{Acceptance Criterion:} When all the players, except one, declare bankruptcy or disconnect permanently, the system declares the remaining player the winner.
    \end{itemize}
\end{itemize}

\subsection{Non-functional requirements}\label{non_functional_requirements}

\begin{itemize}
    \item \textbf{Consistency:} The system should ensure state consistency across all connected nodes at any given time.
    \begin{itemize}
        \item \textit{Acceptance Criterion:} The players GUI's will always show the exact same data.
    \end{itemize}
    \item \textbf{Security:} The system should prevent clients from doing invalid game actions.
    \begin{itemize}
        \item \textit{Acceptance Criterion:} Any altered or malicious payload sent by a client is rejected by the server without affecting the game state.
    \end{itemize}
    \item \textbf{Availability and Fault Tolerance:} The system should be able to recover its exact previous state following a critical node failure.
    \begin{itemize}
        \item \textit{Acceptance Criterion:} Restarting the server after a simulated crash allows clients to reconnect and find the board, balances, and turn order exactly as they were before the crash.
    \end{itemize}
\end{itemize}

\subsection{Implementation requirements}\label{implementation_requirements}

\begin{itemize}
    \item \textbf{Programming Language:} The system is developed using Python.
    \item \textbf{Networking Protocol:} The communication between client and server is
                                              implemented using TCP sockets to ensure reliable delivery.
    \item \textbf{Data Format:} Nodes exchange information using the JSON format.
    \item \textbf{GUI Library:} The client's graphical interface is built using the PyGame library.
    \item \textbf{Containerization:} The server can be started using Docker Compose to facilitate
                                            automatic restarts and environment consistency.
\end{itemize}

\subsection{Relevant Distributed System Features}
\label{ds-features}

\begin{itemize}
  \item \textbf{Transparency:}
  The clients only need a single network endpoint (IP and port) to interact with the service,
  the internal server hosting details are hidden.
  However, full transparency (such as failure or replication transparency) is not a primary goal;
  failures and reconnection states are explicitly made visible to the players via the GUI to manage expectations during network drops.

  \item \textbf{Fault Tolerance and Dependability:}
  If a client fails, recovery systems try to reestablish all the connection and resume the game.
  The server needs to be active for the entire game, but, if a failure occurs,
  it should be able to reload the game state from the last saved checkpoint.
  The checkpoint must be available and not corrupted in order to be used in a server reconnection.
  The reconnections timeouts are 30 seconds for the clients and 90 seconds for the server.
  When a client fails, the reconnection timeout should restart also for the other disconnected players.

  \item \textbf{Scalability:}
  At the moment a server can handle one single game concurrently.
  The system could need to handle an increasing number of users when more lobbies can be created on a single server.
  The amount of data handled should be more or less the same even if the complexity of the game increases as the actual real game is.
  Each game is independent, so at the end of the game everything is reset.


  \item \textbf{Security and Trust:}
  The only data stored are the checkpoint on server's side and the token file for authentication on client's side.
  Cheating is prevented by having a centralized server, clients can only read the serialized game state and cannot communicate to other clients.
  Every player has the same permission.
  The authentication is needed for guarantee that a player that reconnects to a game after
  a failure is the exact same player that was playing before the failure occurred.

  \item \textbf{Resource Sharing:}
  The central shared resource is the global \textit{Game State}.
  The server coordinates and synchronize the process, it handles the incoming events from clients,
  updates the game state and broadcasts it to the connected peers.

  \item \textbf{Communication Efficiency and Bandwidth:}
  The game is turn-based, there is no need to have an instant update from the server.
  All the data sent on the network has almost the same size, from the server is the game state and from the clients the events they generate.
\end{itemize}

\section{Design}\label{design}

This chapter explains the strategies used to meet the requirements identified in the analysis. 
%
Ideally, the design should be the same, regardless of the technological choices made during the implementation phase.

\begin{quote}
You can re-arrange the sections as you prefer, but all the sections must be present in the end
\end{quote}

\noindent
\textbf{Important:} try to motivate your design choices in relation to the requirements and features identified 
in \cref{requirements} and \cref{ds-features}.

\subsection{Architecture}\label{architecture}

\begin{itemize}
  \item Which architectural style?
  %
  \begin{itemize}
    \item why?
  \end{itemize}
\end{itemize}

\subsection{Infrastructure}\label{infrastructure}

\begin{itemize}
  \item are there \emph{infrastructural components} that need to be introduced? \emph{how many}?
  %
  \begin{itemize}
    \item e.g.~\emph{clients}, \emph{servers}, \emph{load balancers},
    \emph{caches}, \emph{databases}, \emph{message brokers},
    \emph{queues}, \emph{workers}, \emph{proxies}, \emph{firewalls},
    \emph{CDNs}, \emph{etc.}
  \end{itemize}

  \item how do components \emph{distribute} over the network? \emph{where}?
  %
  \begin{itemize}
    \item e.g.~do servers / brokers / databases / etc. sit on the same
    machine? on the same network? on the same datacenter? on the same
    continent?
  \end{itemize}

  \item how do components \emph{find} each other?
  %
  \begin{itemize}
    \item how to \emph{name} components?
    \item e.g.~DNS, \emph{service discovery}, \emph{load balancing},
    \emph{etc.}
  \end{itemize}
\end{itemize}

\begin{quote}
Component diagrams are welcome here
\end{quote}

\subsection{Modelling}\label{modelling}

\begin{itemize}
  \item which \textbf{domain entities} are there?
  %
  \begin{itemize}
    \item e.g.~\emph{users}, \emph{products}, \emph{orders}, \emph{etc.}
  \end{itemize}
  
  \item how do \emph{domain entities} \textbf{map to} \emph{infrastructural components}?
  %
  \begin{itemize}
    \item e.g.~state of a video game on central server, while inputs/representations on clients
    \item e.g.~where to store messages in an IM app? for how long?
  \end{itemize}

  \item which \textbf{domain events} are there?
  %
  \begin{itemize}
    \item e.g.~\emph{user registered}, \emph{product added to cart}, \emph{order placed}, \emph{etc.}
  \end{itemize}
  
  \item which sorts of \textbf{messages} are exchanged?
  %
  \begin{itemize}
    \item e.g.~\emph{commands}, \emph{events}, \emph{queries}, \emph{etc.}
  \end{itemize}

  \item what information does the \textbf{state} of the system comprehend
  %
  \begin{itemize}
    \item e.g.~\emph{users' data}, \emph{products' data}, \emph{orders' data}, \emph{etc.}
  \end{itemize}
\end{itemize}

\begin{quote}
Class diagram are welcome here
\end{quote}

\subsection{Interaction}\label{interaction}

\begin{itemize}
  \item how do components \emph{communicate}? \emph{when}? \emph{what}?
  \item \emph{which} \textbf{interaction patterns} do they enact?
\end{itemize}

\begin{quote}
Sequence diagrams are welcome here
\end{quote}

\subsection{Behaviour}\label{behaviour}

\begin{itemize}
  \item how does \emph{each} component \textbf{behave} individually (e.g.~in \emph{response} to \emph{events} or messages)?
  \begin{itemize}
    \item some components may be \emph{stateful}, others \emph{stateless}
  \end{itemize}

  \item which components are in charge of updating the \textbf{state} of the system? \emph{when}? \emph{how}?
\end{itemize}

\begin{quote}
State diagrams are welcome here
\end{quote}

\subsection{Data and Consistency Issues}\label{data-and-consistency-issues}

\begin{itemize}
  \item Is there any data that needs to be stored?
  %
  \begin{itemize}
    \item \emph{what} data? \emph{where}? \emph{why}?
  \end{itemize}
  
  \item how should \emph{persistent data} be \textbf{stored}?
  %
  \begin{itemize}
    \item e.g.~relations, documents, key-value, graph, etc.
    \item why?
  \end{itemize}
  
  \item Which components perform queries on the database?
  %
  \begin{itemize}
    \item \emph{when}? \emph{which} queries? \emph{why}?
    \item concurrent read? concurrent write? why?
  \end{itemize}
  
  \item Is there any data that needs to be shared between components?
  %
  \begin{itemize}
    \item \emph{why}? \emph{what} data?
  \end{itemize}
\end{itemize}

\subsection{Fault-Tolerance}\label{fault-tolerance}

\begin{itemize}
  \item Is there any form of data \textbf{replication} / federation / sharing?
  %
  \begin{itemize}
    \item \emph{why}? \emph{how} does it work?
  \end{itemize}
  
  \item Is there any \textbf{heart-beating}, \textbf{timeout}, \textbf{retry mechanism}?
  %
  \begin{itemize}
    \item \emph{why}? \emph{among} which components? \emph{how} does it work?
  \end{itemize}
  
  \item Is there any form of \textbf{error handling}?
  %
  \begin{itemize}
    \item \emph{what} happens when a component fails? \emph{why}? \emph{how}?
  \end{itemize}
\end{itemize}

\subsection{Availability}\label{availability}

\begin{itemize}
  \item Is there any \textbf{caching} mechanism?
  %
  \begin{itemize}
    \item \emph{where}? \emph{why}?
  \end{itemize}
  
  \item Is there any form of \textbf{load balancing}?
  %
  \begin{itemize}
    \item \emph{where}? \emph{why}?
  \end{itemize}

  \item In case of \textbf{network partitioning}, how does the system behave?
  %
  \begin{itemize}
    \item \emph{why}? \emph{how}?
  \end{itemize}
\end{itemize}

\subsection{Security}\label{security}

\begin{itemize}
  \item Is there any form of \textbf{authentication}?
  %
  \begin{itemize}
    \item \emph{where}? \emph{why}?
  \end{itemize}

  \item Is there any form of \textbf{authorization}?
  %
  \begin{itemize}
    \item which sort of \emph{access control}?
    \item which sorts of users / \emph{roles}? which \emph{access rights}?
  \end{itemize}

  \item Are \textbf{cryptographic schemas} being used?
  %
  \begin{itemize}
    \item e.g.~token verification,
    \item e.g.~data encryption, etc.
  \end{itemize}
\end{itemize}

\section{Implementation}\label{implementation}

Please report here all the implementation (technology-dependent) choices you made while implementing your design.

\begin{quote}
  If you run out of time for this project, you may consider leaving some aspect unimplemented, 
  and simply discuss how you would have implemented them.
  %
  In this case, better would be to discuss unimplemented features in \cref{future-works}.
\end{quote}

\begin{itemize}
  \item which \textbf{network protocols} to use?
  %
  \begin{itemize}
    \item e.g.~UDP, TCP, HTTP, WebSockets, gRPC, XMPP, AMQP, MQTT, etc.
  \end{itemize}
  
  \item how should \emph{in-transit data} be \textbf{represented}?
  %
  \begin{itemize}
    \item e.g.~JSON, XML, YAML, Protocol Buffers, etc.
  \end{itemize}
  
  \item how should \emph{databases} be \textbf{queried}?
  %
  \begin{itemize}
    \item e.g.~SQL, NoSQL, etc.
  \end{itemize}
  
  \item how should components be \emph{authenticated}?
  %
  \begin{itemize}
    \item e.g.~OAuth, JWT, etc.
  \end{itemize}

  \item how should components be \emph{authorized}?
  %
  \begin{itemize}
    \item e.g.~RBAC, ABAC, etc.
  \end{itemize}
\end{itemize}

\subsection{Technological details}\label{technological-details}

\begin{itemize}
  \item any particular \emph{framework} / \emph{technology} being exploited goes here
\end{itemize}

\section{Validation}\label{validation}

\subsection{Automatic Testing}\label{automatic-testing}

\begin{itemize}
  \item how were individual components \textbf{\emph{unit}-test}ed?
  
  \item how was communication, interaction, and/or integration among components tested?
  
  \item how to \textbf{\emph{end-to-end}-test} the system?
  %
  \begin{itemize}
    \item e.g.~production vs.~test environment
  \end{itemize}

  \item for each test specify:
  %
  \begin{itemize}
    \item rationale of individual tests
    \item how were the test automated
    \item how to run them
    \item which requirement they are testing, if any
  \end{itemize}
\end{itemize}

\begin{quote}
recall that \emph{deployment} \textbf{automation} is commonly used to \emph{test} the system in \emph{production-like} environment
\end{quote}

\begin{quote}
recall to test corner cases (crashes, errors, etc.)
\end{quote}

\subsection{Acceptance test}\label{acceptance-test}

\begin{itemize}
  \item did you perform any \emph{manual} testing?
  %
  \begin{itemize}
    \item what did you test?
    \item why wasn't it automatic?
  \end{itemize}
\end{itemize}

\section{Deployment}\label{deployment}

\begin{itemize}
  \item should one install your software from scratch, how to do it?
  %
  \begin{itemize}
    \item provide instructions
    \item provide expected outcomes
  \end{itemize}

  \item what software should be installed on the machines to run your project? which versions?
  
  \item should one set environment variables or configuration files?
  
  \item if you're using containerization (e.g.~Docker), 
  describe how you engineered your deployment scrips (e.g. \texttt{docker-compose.yml} files)
\end{itemize}

\section{User Guide}\label{user-guide}

\begin{itemize}
  \item how to use your software?
  %
  \begin{itemize}
    \item provide instructions
    \item provide expected outcomes
    \item provide screenshots if possible
  \end{itemize}
\end{itemize}

\section{Self-evaluation}\label{self-evaluation}

\begin{itemize}
  \item An individual section is required for each member of the group
  \item Each member must self-evaluate their work, listing the strengths and weaknesses of the product
  \item Each member must describe their role within the group as objectively as possible.
\end{itemize}

It should be noted that each student is only responsible for their own section.

\section{Future works}\label{future-works}

Discuss possible future works, improvements, extensions, optimizations, etc.

Also discuss here any unimplemented feature, and how you would have implemented it.

\bibliographystyle{plain}
\bibliography{references}

\end{document}
